---
title: "Exercise 9"
author: "Kyzer Gerez"
date: "10.12.2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Advanced Methods for Regression and Classification
---

(1) Construct a B-spline basis for \texttt{speed} by using the function 
\texttt{bs()} from the package \texttt{splines}, and fit a linear model. Select 
the degrees of freedom in a way to obtain a plausible fit.
    
    ```{r, echo=TRUE}
    library(splines)
    library(datasets)
    data("cars")
    
    # fit linear model using bs() then predict
    df <- 6
    model.bs <- lm(dist ~ bs(speed, df=df), 
                  data=cars)
    pred.bs <- predict(model.bs, newdata=cars)
    
    # plotting
    plot(cars$speed, cars$dist, col="blue", type="p",
         main=paste("B-spline basis with df=", df, sep=""),
         xlab="Speed",
         ylab="Distance")
    lines(cars$speed, pred.bs)
    ```
    
(2) Construct B-splines for a speed range [0, 30] by using the knots from (1), 
and use the linear model to obtain predictions for the new domain.

    ```{r, echo=TRUE}
    # predict on [0,30]
    x.new <- data.frame(speed = seq(0, 30, length.out = 31))
    pred.bs.ext <- predict(model.bs, newdata=x.new)
    ```
    
(3) Plot the data \texttt{dist} versus \texttt{speed} by extending the plot 
coordinates appropriately, and show the prediction from (2) as a line. Does the 
prediction fulfill the requirements stated above?

    ```{r, echo=TRUE}
    # plotting
    plot(cars$speed, cars$dist, col="blue", type="p",
         main=paste("B-spline basis with df=", df, sep=""),
         xlab="Speed",
         ylab="Distance",
         xlim=c(0,30),
         ylim=c(0, 300))
    lines(x.new$speed, pred.bs.ext)
    ```

    The prediction does not fulfill the requirements. The model predicted 
    negative values for speeds 0 and 1.

(4) Same as (1)-(3), but for \texttt{ns()}.

    ```{r, echo=TRUE}
    # fit linear model using bs() then predict
    df.ns <- 5
    model.ns <- lm(dist ~ ns(speed, df=df.ns), 
                  data=cars)
    pred.ns <- predict(model.bs, newdata=cars)
    
    # predict on [0,30]
    pred.ns.ext <- predict(model.ns, newdata=x.new)
    
    # plotting
    plot(cars$speed, cars$dist, col="blue", type="p",
         main=paste("Natural Cubic Splines with df=", df.ns, sep=""),
         xlab="Speed",
         ylab="Distance",
         xlim=c(0,30),
         ylim=c(0, 200))
    lines(x.new$speed, pred.ns.ext)
    ```
    
    The predictions using the model created from \texttt{ns()} are better than
    those from \texttt{bs()} in that there are no negative predictions. However,
    the predicted distance at speed=0 is still non-zero, though it is a small
    value.
    
    Additionally, the extrapolated values for the high end are less reasonable
    compared to the model created from \texttt{bs()}. There shouldn't be a 
    linear relationship between stopping distance and speed.
    
(5) Use smoothing splines as implemented in \texttt{smooth.spline()}, and select 
the optimal tuning parameter by the internal cross-validation. What are the 
resulting degrees of freedom?

    ```{r, echo=TRUE}
    smooth.spline.cv <- smooth.spline(cars$speed, cars$dist, cv=TRUE)
    pred.smooth <- predict(smooth.spline.cv, newdata=cars)
    ```

    The equivalent number of degrees of freedom is approximately 
    `r smooth.spline.cv$df`. The smoothing parameter, \texttt{spar}, is 
    `r smooth.spline.cv$spar`.

(6) Take the model from (5) to predict for the extended range [0, 30], and show 
the predictions in a plot. What can we do to fulfill the requirements stated 
above?

    ```{r, echo=TRUE}
    # predict on [0,30]
    pred.smooth.ext <- predict(smooth.spline.cv, x=x.new)
    
    # plotting
    plot(cars$speed, cars$dist, col="blue", type="p",
         main="Smoothing Splines",
         xlab="Speed",
         ylab="Distance",
         xlim=c(0,30),
         ylim=c(0, 200))
    lines(pred.smooth.ext$x$speed, pred.smooth.ext$y$speed)
    ```
    
    ```{r, echo=TRUE}
    # use log transform and force 0 dist at speed=0 to meet requirements
    model.smooth <- smooth.spline(cars$speed, log(cars$dist),
                               spar=smooth.spline.cv$spar,
                               df=smooth.spline.cv$df)
    pred.smooth.force <- predict(model.smooth, x=x.new)
    pred.smooth.force$y$speed <- exp(pred.smooth.force$y$speed)
    pred.smooth.force$y$speed[1] <- 0
    
    # plotting
    plot(cars$speed, cars$dist, col="blue", type="p",
         main="Smoothing Splines with Log Transform and Forced 0",
         xlab="Speed",
         ylab="Distance",
         xlim=c(0,30),
         ylim=c(0, 200))
    lines(pred.smooth.force$x$speed, pred.smooth.force$y$speed)
    ```
    Positive predictions can be enforced by using a log transform. A simple
    solution to predicting a stopping distance of 0 at speed=0 is to manually 
    set the prediction to 0 at that speed.
    
    The extrapolation is now reasonable at the high end, but there's a sudden 
    drop in stopping distance from `r pred.smooth.force$y$speed[2]` at speed=1
    to 0 at speed=0.

</div></pre>