---
title: "Exercise 6"
author: "Kyzer Gerez"
date: "19.11.2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Advanced Methods for Regression and Classification
---

```{r, echo=TRUE}
library(fastDummies)

# read csv and drop duration col
d <- subset(read.csv2("bank.csv"), select=-c(duration))

# check for na
rownames(d[rowSums(is.na(d)) > 0,])

# convert month to numeric representation
d["month"] <- match(d[["month"]], tolower(month.abb))

# convert binary variables to numeric representation
d$default <- ifelse(d$default=="no", 0, 1)
d$housing <- ifelse(d$housing=="no", 0, 1)
d$loan <- ifelse(d$loan=="no", 0, 1)
d$y <- ifelse(d$y=="no", 0, 1)

# apply one-hot encoding to rest of categorical predictors
d.encoding <- dummy_cols(d, 
                         remove_selected_columns=TRUE)

# split into 3/4 train, 1/4 test
smp_size <- floor(3/4 * nrow(d))

# set the seed to make partition reproducible
set.seed(1)
train_ind <- sample(seq_len(nrow(d.encoding)), size = smp_size)

train <- d.encoding[train_ind, ]
test <- d.encoding[-train_ind, ]
```

1. \textit{Linear Discrimant Analysis (LDA)}: function \texttt{lda} from 
\texttt{library(MASS)}

    (a) Apply \texttt{lda()} to the training set. Is any data preprocessing 
    necessary or advisable?
    
    Yes, data preprocessing is necessary since LDA requires that all predictors 
    be in numeric form and scaled. One hot-encoding should be used for nominal
    predictors such as job and marital, while the binary predictors and ordinal 
    month predictor can simply be converted to a numerical form. Education could 
    also be treated as an ordinal variable, but it's unclear how unknown values 
    should be handled; therefore, it's treated as a nominal variable instead.
    
    Additionally, unknown values will be treated as their own category and will
    not be imputed.
    
    ```{r, echo=TRUE}
    library(MASS)
    
    mod.lda <- lda(x = train[,-11], grouping = train$y) # remove y for x param
    ```
    
    (b) Compute the evaluation measures for the training data. What do you conclude?
    
    ```{r echo=TRUE, message=FALSE}
    lda.train.pred <- predict(mod.lda, newdata=train[,-11])
    
    # create confusion matrix
    library(caret)
    
    lda.train.confusion <- confusionMatrix(data=lda.train.pred$class, 
                                           reference=as.factor(train$y),
                                           positive = "1")
    lda.train.confusion
    ```
    
    The model tends to predict no due to the unbalanced dataset. This can be 
    seen from the high accuracy compared to the lower values for sensitivity and
    balanced accuracy.
    
    (c) Predict the group membership for the test data and compute the evaluation
    measures. What do you conclude?
    
    ```{r, echo=TRUE}
    lda.test.pred <- predict(mod.lda, newdata=test[,-11])
    
    lda.test.confusion <- confusionMatrix(data=lda.test.pred$class,
                                           reference=as.factor(test$y),
                                           positive = "1")
    lda.test.confusion
    ```
    
    Again, the same results can be seen in that the model overpredicts for no
    due to the unbalanced dataset. 
    
2. Assume that we want to maximize the balanced accuracy. This could be achieved
by selecting a balanced training set, i.e., the same number of observations from both
groups, to train the classifier. Thus, repeat tasks 1.(a)-(c) for:

    ```{r, echo=TRUE}
    # setup #
    n_0 <- length(which(d.encoding$y==0)) # max
    n_1 <- length(which(d.encoding$y==1)) # min
    
    train.only.0 <- d.encoding[which(d.encoding$y==0),]
    train.only.1 <- d.encoding[which(d.encoding$y==1),]
    
    # get undersampled train set
    under.indices <- sample(seq_len(nrow(train.only.0)), n_1)
    under.0 <- train.only.0[under.indices,]
    train.under <- rbind(train.only.1, 
                         under.0)
    
    # get oversampled train set
    over.indices <- sample(seq_len(nrow(train.only.1)), n_0, replace=TRUE)
    over.1 <- train.only.1[over.indices,]
    train.over <- rbind(train.only.0, 
                         over.1)
    ```

    (a) “Undersampling”: Suppose the number of samples in the original training 
    set is $n_1$ and $n_2$ for the two groups. Now select from this training set 
    min($n_1$, $n_2$) samples from each group. The test set is unchanged.
    
    ```{r, echo=TRUE}
    # train model
    mod.lda.under <- lda(x = train.under[,-11], grouping = train.under$y)
    
    # predict on train and eval
    lda.under.train.pred <- predict(mod.lda.under, newdata=train[,-11])
    lda.under.train.confusion <- confusionMatrix(data=lda.under.train.pred$class,
                                                 reference=as.factor(train$y),
                                                 positive = "1")
    lda.under.train.confusion
    
    # predict on test and eval
    lda.under.test.pred <- predict(mod.lda.under, newdata=test[,-11])
    lda.under.test.confusion <- confusionMatrix(data=lda.under.test.pred$class,
                                                 reference=as.factor(test$y),
                                                 positive = "1")
    lda.under.test.confusion
    ```
    
    (b) “Oversampling”: Select from the training set max($n_1$, $n_2$) samples 
    from each group. From the smaller group you need to sample with replacement. 
    The test set is unchanged.
    
    ```{r, echo=TRUE}
    # train model
    mod.lda.over <- lda(x = train.over[,-11], grouping = train.over$y)
    
    # predict on train and eval
    lda.over.train.pred <- predict(mod.lda.over, newdata=train[,-11])
    lda.over.train.confusion <- confusionMatrix(data=lda.over.train.pred$class,
                                                 reference=as.factor(train$y),
                                                 positive = "1")
    lda.over.train.confusion
    
    # predict on test and eval
    lda.over.test.pred <- predict(mod.lda.over, newdata=test[,-11])
    lda.over.test.confusion <- confusionMatrix(data=lda.over.test.pred$class,
                                                 reference=as.factor(test$y),
                                                 positive = "1")
    lda.over.test.confusion
    ```
    
    Which strategy is more successful, and why?
    
    The oversampling strategy performed slightly better for the test set, while
    undersampling resulted in a higher balanced accuracy for the train set. This 
    is a bit unexpected since oversampling should lead to more overfitting. But
    it being better for the test set makes sense. The disadvantage of 
    undersampling is that information from the majority class is discarded, 
    which can lead to weaker predictive power.

3. \textit{Quadratic Discrimant Analysis (QDA)}: function \texttt{qda} from
\texttt{library(MASS)}. Use oversampling and undersampling for \texttt{qda()}
and report the evaluation measures for the test set. What do you conclude?

    ```{r, echo=TRUE}
    library(MASS)
    
    ### QDA-specific preprocessing ###
    # check if class 0 predictors are full rank
    X0 <- as.matrix(train[train$y == 0, -which(names(train) == "y")])
    k <- qr(X0)$rank
    p <- ncol(X0)
    k; p
    print("k<p, so class 0 predictors are not full rank.")
    
    # check for linear combos
    remove.cols <- findLinearCombos(X0)$remove
    
    # remove predictors 25 29 32 36
    train.qda <- train[, -remove.cols]
    train.qda.under <- train.under[, -remove.cols]
    train.qda.over <- train.over[, -remove.cols]
    test.qda <- test[, -remove.cols]
    
    # find columns with near-zero variance
    nzv_cols <- nearZeroVar(X0, saveMetrics = TRUE)
    
    # remove columns with near-zero variance
    train.qda <- train.qda[, !names(train.qda) %in% 
                             rownames(nzv_cols[nzv_cols$nzv == TRUE, ])]
    train.qda.under <- train.qda.under[, !names(train.qda.under) %in% 
                             rownames(nzv_cols[nzv_cols$nzv == TRUE, ])]
    train.qda.over <- train.qda.over[, !names(train.qda.over) %in% 
                             rownames(nzv_cols[nzv_cols$nzv == TRUE, ])]
    test.qda <- test.qda[, !names(test.qda) %in% 
                             rownames(nzv_cols[nzv_cols$nzv == TRUE, ])]
    
    ### undersampling ###
    # train model
    mod.qda.under <- qda(y ~ ., data=train.qda.under)
    
    # predict on train and eval
    qda.under.train.pred <- predict(mod.qda.under, newdata=train.qda[,-9])
    qda.under.train.confusion <- confusionMatrix(data=qda.under.train.pred$class,
                                                 reference=as.factor(train.qda$y),
                                                 positive = "1")
    qda.under.train.confusion
    # predict on test and eval
    qda.under.test.pred <- predict(mod.qda.under, newdata=test.qda[,-9])
    qda.under.test.confusion <- confusionMatrix(data=qda.under.test.pred$class,
                                                 reference=as.factor(test.qda$y),
                                                 positive = "1")
    qda.under.test.confusion
    
    ## oversampling ##
    # train model
    mod.qda.over <- qda(y ~ ., data=train.qda.over)
    
    # predict on train and eval
    qda.over.train.pred <- predict(mod.qda.over, newdata=train.qda[,-9])
    qda.over.train.confusion <- confusionMatrix(data=qda.over.train.pred$class,
                                                 reference=as.factor(train.qda$y),
                                                 positive = "1")
    qda.over.train.confusion
    # predict on test and eval
    qda.over.test.pred <- predict(mod.qda.over, newdata=test.qda[,-9])
    qda.over.test.confusion <- confusionMatrix(data=qda.over.test.pred$class,
                                                 reference=as.factor(test.qda$y),
                                                 positive = "1")
    qda.over.test.confusion
    ```
    
    Significant additional data-preprocessing was needed for QDA to work. QDA is
    brittle in that it can't handle predictors that have near zero variance or
    are linear combinations of other predictors.
    
    The oversampling strategy outperformed for both the train and test sets.

4. \textit{Regularized Discrimant Analysis (RDA)}: function \texttt{rda} from
\texttt{library(klaR)} Use oversampling and undersampling for \texttt{rda()} and 
report the evaluation measures for the test set. What do you conclude? Interpret the meaning of the resulting tuning parameters \texttt{gamma} and \texttt{lambda}.

    ```{r, echo=TRUE}
    library(klaR)
    
    ## undersampling ##
    # train model
    mod.rda.under <- rda(x = train.under[,-11], grouping = train.under$y)
    
    # predict on train and eval
    rda.under.train.pred <- predict(mod.rda.under, newdata=train[,-11])
    rda.under.train.confusion <- confusionMatrix(data=rda.under.train.pred$class,
                                                 reference=as.factor(train$y),
                                                 positive = "1")
    rda.under.train.confusion
    
    # predict on test and eval
    rda.under.test.pred <- predict(mod.rda.under, newdata=test[,-11])
    rda.under.test.confusion <- confusionMatrix(data=rda.under.test.pred$class,
                                                 reference=as.factor(test$y),
                                                 positive = "1")
    rda.under.test.confusion
    
    ## oversampling ##
    # train model
    mod.rda.over <- rda(x = train.over[,-11], grouping = train.over$y)
    
    # predict on train and eval
    rda.over.train.pred <- predict(mod.rda.over, newdata=train[,-11])
    rda.over.train.confusion <- confusionMatrix(data=rda.over.train.pred$class,
                                                 reference=as.factor(train$y),
                                                 positive = "1")
    rda.over.train.confusion
    
    # predict on test and eval
    rda.over.test.pred <- predict(mod.rda.over, newdata=test[,-11])
    rda.over.test.confusion <- confusionMatrix(data=rda.over.test.pred$class,
                                                 reference=as.factor(test$y),
                                                 positive = "1")
    rda.over.test.confusion
    
    # output tuning parameters
    rda.tuning.params <- data.frame(
      model = c("under", "over"),
      gamma = c(mod.rda.under$regularization[["gamma"]],
                mod.rda.over$regularization[["gamma"]]),
      lambda = c(mod.rda.under$regularization[["lambda"]],
                mod.rda.over$regularization[["lambda"]])
    )
    rda.tuning.params
    ```
The oversampling strategy performed significantly better than undersampling for
both the train and test sets. RDA is a compromise between QDA and LDA. The lambda
values for the two sampling strategies indicate that the estimated error rate is
minimized using values that shape the models to being closer to LDA than QDA.
The high values for gamma for both sampling strategies show that the covariance
structure was shrunk to a more spherical shape. This means that predictors are
more equally weighted compared to pure QDA and LDA.

```{r, echo=TRUE}
# summary of results
summary <- data.frame(
  model = c("lda", "lda.under", "lda.over", "qda.under", "qda.over", 
            "rda.under", "rda.over"),
  train_balanced_acc = c(lda.train.confusion$byClass[["Balanced Accuracy"]],
                         lda.under.train.confusion$byClass[["Balanced Accuracy"]],
                         lda.over.train.confusion$byClass[["Balanced Accuracy"]],
                         qda.under.train.confusion$byClass[["Balanced Accuracy"]],
                         qda.over.train.confusion$byClass[["Balanced Accuracy"]],
                         rda.under.train.confusion$byClass[["Balanced Accuracy"]],
                         rda.over.train.confusion$byClass[["Balanced Accuracy"]]),
  test_balanced_acc = c(lda.test.confusion$byClass[["Balanced Accuracy"]],
                        lda.under.test.confusion$byClass[["Balanced Accuracy"]],
                        lda.over.test.confusion$byClass[["Balanced Accuracy"]],
                        qda.under.test.confusion$byClass[["Balanced Accuracy"]],
                        qda.over.test.confusion$byClass[["Balanced Accuracy"]],
                        rda.under.test.confusion$byClass[["Balanced Accuracy"]],
                        rda.over.test.confusion$byClass[["Balanced Accuracy"]])
  )
summary
```

Overall, QDA using the oversampling strategy outperformed all other models for
both the train and test sets. Though comparisons between QDA and the other
methods aren't on a completely sound basis since QDA required train and test
sets with less predictors than the sets used to create the LDA and RDA models.

</div></pre>