---
title: "Exercise 11"
author: "Kyzer Gerez"
date: "14.01.2026"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Advanced Methods for Regression and Classification
---

Take the data set \texttt{german.credit} from the package \texttt{fairml} 
(see help page). The variable \texttt{Credit\_risk} is the grouping variable, 
and it indicates whether the persons were successful with their loan application 
or not. The "BAD" clients are particularly interesting for us, but they are in 
the minority (= imbalanced classes)

For the following tasks, select randomly about 2/3 of the observations from each class
as training data, and the rest as test data.

```{r echo=TRUE, message=FALSE}
library(fairml); data(german.credit)
library(rpart)
library(rpart.plot)
library(randomForest)
library(caret)

set.seed(12427543)

# separate the two classes
credit.BAD <- german.credit[german.credit$Credit_risk == 'BAD',]
credit.GOOD <- german.credit[german.credit$Credit_risk == 'GOOD',]

# get sample sizes for each class
samp.size.BAD <- floor(2/3*nrow(credit.BAD))
samp.size.GOOD <- floor(2/3*nrow(credit.GOOD))

# create train index
train.indx <- sample(as.numeric(rownames(credit.BAD)), size=samp.size.BAD)
train.indx <- c(train.indx, sample(as.numeric(rownames(credit.GOOD)), samp.size.GOOD))

# create stratified train and test splits
train <- german.credit[train.indx,]
test <- german.credit[-train.indx,]
```

1. *Classification trees*: function \texttt{rpart()} from the R package 
\texttt{rpart}
    (a) Compute an initial tree $T_0$ (see \texttt{help(rpart)} or lecture notes).
    
    ```{r, echo=TRUE}
    mod.T_0 <- rpart(Credit_risk~., data=train, method='class', xval=20)
    ```

    (b) Visualize the tree with the function \texttt{rpart.plot()} from the 
    package \texttt{rpart.plot}, and try to interpret the results.
    
    ```{r, echo=TRUE}
    rpart.plot(mod.T_0)
    ```
    
    Each node shows the predicted class and the percentage of observations in 
    the node. At each branch, if the conditions are met, then take the left 
    subtree. Else, take the right subtree. The middle number in each node 
    represents the probability that an observation that falls to this node has
    a good credit risk i.e. nodes labelled 'BAD' will have low values.
    
    For the above tree, the first branch checks the account status. If it's less
    than 200 DM, the left subtree is taken. At this next level, 56% of 
    observations go to the left subtree, while the rest take the right. This
    process repeats until a leaf node is reached. 
    
    (c) Predict the class variable for the test set (see 
    \texttt{help(predict.rpart)} or lecture notes). Show the confusion table and 
    report the balanced accuracy.
    
    ```{r, echo=TRUE}
    # predict
    pred.T_0 <- predict(mod.T_0, newdata=test, type='class')
    
    # create confusion matrix and report required output
    confusionMatrix.T_0 <- confusionMatrix(as.factor(pred.T_0), 
                                           as.factor(test$Credit_risk))
    confusionMatrix.T_0$table
    confusionMatrix.T_0$byClass['Balanced Accuracy']
    ```
    
    (d) Show and interpret results of cross-validation obtained by using 
    \texttt{plotcp()}. What is the optimal tree complexity?
    
    ```{r, echo=TRUE}
    printcp(mod.T_0)
    ```
    
    ```{r, echo=TRUE}
    plotcp(mod.T_0)
    ```
    
    The optimal tree complexity appears to be 0.0125.
    
    (e) Prune the tree $T_0$ to the optimal complexity using \texttt{prune()}. 
    Visualize and interpret the results.
    
    ```{r, echo=TRUE}
    mod.T_0.pruned <- prune(mod.T_0, cp=0.0125, method='class')
    
    rpart.plot(mod.T_0.pruned)
    ```
    
    The pruned tree is of course simpler than the initial tree. The same 
    variables are used. However, there are only 7 splits compared to 12 for the
    initial tree.
    
    (f) Predict the class variable for the test set, show the confusion table, 
    and report the balanced accuracy. Do we observe any improvement?
    
    ```{r, echo=TRUE}
    # predict
    pred.T_0.pruned <- predict(mod.T_0.pruned, newdata=test, type='class')
    
    # create confusion matrix and report required output
    confusionMatrix.T_0.pruned <- confusionMatrix(as.factor(pred.T_0.pruned),
                                                  as.factor(test$Credit_risk))
    confusionMatrix.T_0.pruned$table
    confusionMatrix.T_0.pruned$byClass['Balanced Accuracy']
    ```

    Overall, the pruned tree led to worse results in terms of the balanced 
    accuracy. It was better in predicting the "GOOD" class, but performed much
    worse in predicting the 'BAD' class, which led to a lower balanced accuracy.

    (g) A simple way to improve the balanced accuracy could be to make use of the
    argument \texttt{weights} within \texttt{rpart()}. Try it out and report the 
    results.
    
    The approach I will take is to find the proportions of each class and use
    the proportion for the opposite class as the weight.
    
    ```{r, echo=TRUE}
    # create a weight vector based on the Credit_risk values
    weight.BAD <- nrow(credit.GOOD)/nrow(german.credit)
    weight.GOOD <- nrow(credit.BAD)/nrow(german.credit)
    weights_vec <- ifelse(train$Credit_risk == "BAD", weight.BAD, weight.GOOD)
    
    # create weighted tree
    mod.T.weight <- rpart(Credit_risk~ ., data=train, weights=weights_vec)
    
    # predict
    pred.T.weight <- predict(mod.T.weight, newdata=test, type='class')
    
    # create confusion matrix and report required output
    confusionMatrix.T.weight <- confusionMatrix(as.factor(pred.T.weight),
                                                as.factor(test$Credit_risk))
    confusionMatrix.T.weight$table
    confusionMatrix.T.weight$byClass['Balanced Accuracy']
    ```

    The weights used led to an improved balanced accuracy. The 
    prediction for the "BAD" class improved at the cost of the "GOOD" class. 
    But this is tolerable since the "BAD" clients are the ones of interest.
    
1. *Random Forests*: function \texttt{randomForest()} from the R package 
\texttt{randomForest}

    (a) Use Random Forests to classify the training data and predict the class 
    variable for the test data. Report the resulting confusion table and the 
    balanced accuracy.
    
    ```{r, echo=TRUE}
    # train random forest on default parameters
    mod.RF_0 <- randomForest(Credit_risk~ ., data=train)
    
    # predict
    pred.RF_0 <- predict(mod.RF_0, test, type='class')
    
    # create confusion matrix and report required output
    confusionMatrix.RF_0 <- confusionMatrix(as.factor(pred.RF_0),
                                            as.factor(test$Credit_risk))
    confusionMatrix.RF_0$table
    confusionMatrix.RF_0$byClass['Balanced Accuracy']
    ```
    
    (b) Plot the result object with \texttt{plot()} and interpret the plot.
    
    ```{r, echo=TRUE}
    plot(mod.RF_0)
    legend("right", 
           legend = c("OOB", levels(train$Credit_risk)), 
           col = c("black", "red", "green"), 
           lty = 1, 
           cex = 0.8)
    ```
    
    ```{r, echo=TRUE}
    mod.RF_0[["err.rate"]][1:10,]
    ```
    
    The y-axis is error rates and x-axis is the number of trees. The black line
    represents the overall error rate, the red line is the error rate for the
    "BAD" class, and the green line is the error rate for the "GOOD" class.
    
    The plot shows that the overall and "GOOD" error rates decrease with an 
    increasing number of trees. However, the opposite behavior can be seen with
    for the error rate of the "BAD" class.
    
    An inspection of the error rates show that the minimum error rate for the
    "BAD" class is achieved with only 2 trees.
    
    (c) Try to improve the balanced accuracy with different strategies:
    
    - Modify the parameter \texttt{sampsize} in the \texttt{randomForest()}
    function. What is it doing?
    
        This parameter maps to the bootstrap sample size. This means that it 
        controls how many observations are randomly sampled from each node 
        during the construction of the random forest model. With 
        \texttt{replace=TRUE} (default), \texttt{sampsize=nrow(x)}. 
        
        To determine the value to set the parameter to, I tested different 
        of $x$ for ceiling(x*nrow(train))) in 0.1 increments from 0.1 to
        1.0. The value that maximizes the balanced accuracy is 0.5.
    
    ```{r, echo=TRUE}
    # train RF
    set.seed(12427543)
    mod.RF.sampsize <- randomForest(Credit_risk~ ., data=train,
                                    sampsize=ceiling(0.5*nrow(train)))
    
    # predict
    pred.RF.sampsize <- predict(mod.RF.sampsize, test, type='class')
    
    # create confusion matrix and report required output
    confusionMatrix.RF.sampsize <- confusionMatrix(as.factor(pred.RF.sampsize),
                                                   as.factor(test$Credit_risk))
    confusionMatrix.RF.sampsize$table
    confusionMatrix.RF.sampsize$byClass['Balanced Accuracy']
    ```
    
    - Modify the parameter \texttt{classwt} in the \texttt{randomForest()} 
    function. What is it doing?
        
        This has a similar function as the \texttt{weights} parameter for 
        \texttt{rpart()} in that it can be used to deal with imbalanced 
        classes. In the below approach, I use a similar weight vector as in 
        part 1g.
    
    ```{r, echo=TRUE}
    set.seed(12427543)
    
    # train RF
    mod.RF.classwt <- randomForest(Credit_risk~ ., data=train, 
                                   classwt=c(weight.BAD, weight.GOOD))
    
    # predict
    pred.RF.classwt <- predict(mod.RF.classwt, test, type='class')
    
    # create confusion matrix and report required output
    confusionMatrix.RF.classwt <- confusionMatrix(as.factor(pred.RF.classwt),
                                                  as.factor(test$Credit_risk))
    confusionMatrix.RF.classwt$table
    confusionMatrix.RF.classwt$byClass['Balanced Accuracy']
    ```
    
    - Modify the parameter \texttt{cutoff} in the \texttt{randomForest()} 
    function. What is it doing?
        
        This parameter controls the number of votes needed to determine the class
        of an observation. By default, the majority vote wins i.e. 1/$k$ where 
        $k$ is the number of classes.
        
        For the cell below, I tested different combinations of the cutoff vector
        in increments of 0.05 e.g. \texttt{c(0.05, 0.95), c(0.10, 0.90), ...}
        The best results were achieved with a cutoff vector of 
        \texttt{c(0.35, 0.65)} favoring the "BAD" class.
         
    ```{r, echo=TRUE}
    set.seed(12427543)
    
    # train RF
    mod.RF.cutoff <- randomForest(Credit_risk~ ., data=train, 
                                  cutoff=c(0.35, 0.65))
    
    # predict
    pred.RF.cutoff <- predict(mod.RF.cutoff, test, type='class')
    
    # create confusion matrix and report required output
    confusionMatrix.RF.cutoff <- confusionMatrix(as.factor(pred.RF.cutoff),
                                                  as.factor(test$Credit_risk))
    confusionMatrix.RF.cutoff$table
    confusionMatrix.RF.cutoff$byClass['Balanced Accuracy']
    ```
    
      **Which approach leads to the overall best solution?**
      
      Overall, changing the \texttt{cutoff} parameter gave the best solution
      by far. It resulted in a balanced accuracy of 71.3% compared to 61.9%
      for the default random forest model.
      
      Modfying \texttt{sampsize} led to a marginal improvement. It correctly 
      predicted two more "BAD" observations,which is an increase in the 
      balanced accuracy of 1% (absolute) over the default random forest model.
      
      The weights chosen for the \texttt{classwt} yielded the same results as
      the default random forest model.
    
    (d) Run the function \texttt{randomForest()} based on the best strategy from 
    above, with the additional option \texttt{importance=TRUE}, and plot the 
    result object with \texttt{plot()} and \texttt{varImpPlot()}. How can you 
    interpret these plots?
    
    ```{r, echo=TRUE}
    set.seed(12427543)
    
    # train RF
    mod.RF.best <- randomForest(Credit_risk~ ., data=train, 
                                cutoff=c(0.35, 0.65),
                                importance=TRUE)
    
    plot(mod.RF.best)
    legend("topright", 
           legend = c("OOB", levels(train$Credit_risk)), 
           col = c("black", "red", "green"), 
           lty = 1, 
           cex = 0.8)
    ```
    
    ```{r, echo=TRUE}
    varImpPlot(mod.RF.best, type=1)
    ```
    
    ```{r, echo=TRUE}
    varImpPlot(mod.RF.best, type=2)
    ```

    The results for \texttt{plot()} shows much better behavior than that from
    the default random forest model in that increasing the number of trees
    decreases the error rate for both classes. The error rate for the "GOOD"
    class does settle at a higher value than in the default model, but this
    is acceptable since the performance on the "BAD" class is greatly improved.
    
    The function \texttt{varImpPlot()} displays the importance of variables 
    based on two different measures. The top plot is based on the mean
    decrease in accuracy. The bottom plot uses the mean decrease in node
    impurity.
    
    The lecture notes explain that the Gini index or node impurity is preferred
    over misclassification error Given this recommendation, the top three
    most important variables are \texttt{Credit\_amount}, 
    \texttt{Account\_status}, and \texttt{Duration}.
    
    The only disagreement with the misclassification error method in the top-3
    is that it swaps \texttt{Credit\_amount} for \texttt{Credit\_history}.

</div></pre>