---
title: "Exercise 7"
author: "Kyzer Gerez"
date: "26.11.2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Advanced Methods for Regression and Classification
---

```{r, echo=TRUE}
library(fastDummies)

# read csv and drop duration col
d <- subset(read.csv2("bank.csv"), select=-c(duration))

# check for na
rownames(d[rowSums(is.na(d)) > 0,])

# convert month to numeric representation
d["month"] <- match(d[["month"]], tolower(month.abb))

# convert binary variables to numeric representation
d$default <- ifelse(d$default=="no", 0, 1)
d$housing <- ifelse(d$housing=="no", 0, 1)
d$loan <- ifelse(d$loan=="no", 0, 1)
d$y <- ifelse(d$y=="no", 0, 1)

# split into 3/4 train, 1/4 test
smp_size <- floor(3/4 * nrow(d))

# set the seed to make partition reproducible
set.seed(1)
train_ind <- sample(seq_len(nrow(d)), size = smp_size)

train <- d[train_ind, ]
test <- d[-train_ind, ]
```

1. Use the same data set \texttt{bank.csv} as in the last exercise, also 
without variable “duration”.

    (a) Use logistic regression for a training set (function \texttt{glm()} with
    \texttt{family="binomial"}). Look at the inference table (with 
    \texttt{summary()}) and interpret the outcome.
    
    ```{r, echo=TRUE}
    library(MASS)
    
    # train model
    mod.log.reg <- glm(y ~., train, family="binomial")
    
    summary(mod.log.reg)
    ```
  
    The summary shows that the predictors marital, loan, housing, and 
    campaign are significant in predicting the response. Additionally, the 
    values job-retired, education-tertiary, contact-unkown, and poutcome-success
    are significant.
    
    (b) Use the model to predict the group label of the remaining test set 
    observations (what does the function actually predict by default?). Compute 
    the misclassification rate for every group separately, show the confusion 
    table, and report the balanced accuracy.
    
    The misclassification rate for each group can be calculated as 
    $\frac{FN+FP}{n}$ where $n$ is the number of instances for the given group.
    
    ```{r echo=TRUE, message=FALSE}
    bank.test.pred <- predict(mod.log.reg, test)
    
    plot(bank.test.pred,col= test$y+2,
         ylab="Predicted")
    
    # convert from linear to binary scale
    bank.test.pred.mod <- ifelse(bank.test.pred<=0, 0, 1)
      
    # confusion matrix and balanced accuracy
    library(caret)
    
    bank.test.confusion <- confusionMatrix(data=as.factor(bank.test.pred.mod),
                                           reference=as.factor(test$y),
                                           positive = "1")
    bank.test.confusion$table
    bank.test.confusion$byClass["Balanced Accuracy"]
    
    # compute misclassification rate for no
    no.total <- bank.test.confusion$table[1,1] + bank.test.confusion$table[2,1]
    no.FN <- bank.test.confusion$table[2,1]
    no.FP <- bank.test.confusion$table[1,2]
    bank.misclass.no <- round((no.FN + no.FP) / no.total, 3)
    
    # compute misclassification rate for yes
    yes.total <- bank.test.confusion$table[1,2] + bank.test.confusion$table[2,2]
    yes.FN <- bank.test.confusion$table[1,2]
    yes.FP <- bank.test.confusion$table[2,1]
    bank.misclass.yes <- round((yes.FN + yes.FP) / yes.total, 3)
    
    # output misclassification rates
    print(paste("The misclassification rate for 'no' is", bank.misclass.no))
    print(paste("The misclassification rate for 'yes' is", bank.misclass.yes))
    ```
    
    The predictions are returned in the scale of the linear predictor, with
    zero as the decision boundary. The model predicts "no" by default since the 
    intercept is a negative value.
    
    The colors in the plot indicates the true class of the observation, with red
    being "no" and green being "yes".
    
    (c) A way to consider the problem of imbalanced groups is to assign a weight 
    to every observation, by using the weights argument in the \texttt{glm()} 
    function. How do you have to select the weights, and what is the resulting 
    confusion table and balanced accuracy?
    
    The weights can simply be the proprtion of each class for the train set.
    The weight for the minority class would be the proportion of the majority
    class and vice versa. 
    
    ```{r echo=TRUE, warning=FALSE}
    # calculate weights
    weight.0 <- nrow(train[train$y == 1,])/ nrow(train)
    weight.1 <- nrow(train[train$y == 0,])/ nrow(train)
    
    # create weight vector
    weights <- ifelse(train$y == 0, weight.0, weight.1)
    
    # train weighted model
    mod.weighted <- glm(y ~., train, 
                        family="binomial",
                        weights=weights)
    
    # predict
    bank.weighted.test.pred <- predict(mod.weighted, test)
    
    plot(bank.weighted.test.pred,col= test$y+2,
         ylab="Predicted")
    
    # convert from linear to binary scale
    bank.weighted.test.pred.mod <- ifelse(bank.weighted.test.pred<=0, 0, 1)
    
    # confusion matrix and balanced accuracy
    bank.weighted.test.confusion <- confusionMatrix(data=as.factor(bank.weighted.test.pred.mod),
                                           reference=as.factor(test$y),
                                           positive = "1")
    bank.weighted.test.confusion$table
    bank.weighted.test.confusion$byClass["Balanced Accuracy"]
    
    # compute misclassification rate for no
    no.total <- bank.weighted.test.confusion$table[1,1] +
      bank.weighted.test.confusion$table[2,1]
    no.FN <- bank.weighted.test.confusion$table[2,1]
    no.FP <- bank.weighted.test.confusion$table[1,2]
    bank.misclass.no <- round((no.FN + no.FP) / no.total, 3)
    
    # compute misclassification rate for yes
    yes.total <- bank.weighted.test.confusion$table[1,2] +
      bank.weighted.test.confusion$table[2,2]
    yes.FN <- bank.weighted.test.confusion$table[1,2]
    yes.FP <- bank.weighted.test.confusion$table[2,1]
    bank.misclass.yes <- round((yes.FN + yes.FP) / yes.total, 3)
    
    # output misclassification rates
    print(paste("The misclassification rate for 'no' is", bank.misclass.no))
    print(paste("The misclassification rate for 'yes' is", bank.misclass.yes))
    ```
    
    The weighted model has an improved balanced accuracy at the cost of a much
    higher misclassification rate for the 'yes' group.
    
    (d) Based on the model from 1(c), use stepwise variable selection with the 
    function \texttt{step()} to simplify the model. Does this also lead to an 
    improvement of the balanced accuracy?
    
    ```{r echo=TRUE, warning=FALSE}
    # backward selection
    mod.back <- step(mod.weighted,
                     direction = "backward",
                     trace=0)
    
    # predict
    bank.back.test.pred <- predict(mod.back, test)
    
    plot(bank.back.test.pred,col= test$y+2,
         ylab="Predicted")
    
    # convert from linear to binary scale
    bank.back.test.pred.mod <- ifelse(bank.back.test.pred<=0, 0, 1)
    
    # confusion matrix and balanced accuracy
    bank.back.test.confusion <- confusionMatrix(data=as.factor(bank.back.test.pred.mod),
                                           reference=as.factor(test$y),
                                           positive = "1")
    bank.back.test.confusion$table
    bank.back.test.confusion$byClass["Balanced Accuracy"]
    ```
    
    The balanced accuracy for the backward selection model is 
    `r round(bank.back.test.confusion$byClass["Balanced Accuracy"], 3)`.
    This is slightly better than the balanced accuracy for the original model,
    which is 
    `r round(bank.weighted.test.confusion$byClass["Balanced Accuracy"], 3)`.
    
2. Use the data set \texttt{data(Khan)} from the package \texttt{ISLR}. The data 
set consists of a number of tissue samples corresponding to four distinct types 
of small round blue cell tumors. For each tissue sample, 2308 gene expression 
measurements are available (see also help file). The task is to train a 
classifier based on the training data (\texttt{Khan\$xtrain}, \texttt{Khan\$ytrain}),
use it for predicting the class of the test data (\texttt{Khan\$xtest}), and to 
evaluate the predictions using the group information of the test data 
(\texttt{Khan\$ytest}).

    ```{r, echo=TRUE}
    library(ISLR)
    data(Khan)
    
    print(paste("n =", nrow(Khan$xtrain)))
    print(paste("p =", ncol(Khan$xtrain)))
    ```

    (a) Why would LDA or QDA not work here? Would RDA work (you can either
    try it out, or simply argue)?
    
    LDA and QDA requires that the sample covariance matrix to be full rank
    and invertable, which does not happen if $n<p$. For the Khan dataset, $n<p$.
    In contrast, RDA works under this condition because it ensures stability
    of the covariance matrix by shifting it towards a diagonal matrix or pooled
    covariance matrix using the two tuning parameters.
    
    (b) Use the function \texttt{cv.glmnet()} from the package \texttt{glmnet}, 
    with the argument \texttt{family="multinomial"}, to build a model for the 
    training set (the response might need to be converted to a factor). Plot the 
    outcome object. What do you conclude? What is the objective function to be 
    minimized?
    
    ```{r echo=TRUE, warning=FALSE}
    library(glmnet)
    
    mod.multi.glm <- cv.glmnet(x=Khan$xtrain, y=Khan$ytrain, 
                               family="multinomial")
    plot(mod.multi.glm, sign.lambda=1)
    ```
    
    The plot shows that the best performing models have small lambda values, 
    which means that coefficients are not shrunk significantly. However, the 
    best models still contain only nine non-zero coefficients out of the 
    original 2308 predictors.
    
    The objective function to be minimized is the log-likelihood function, and
    the results are represented by the y-axis, multinomial deviance. The 
    deviance is the negative contribution of each observation to the 
    log-likelihood function.
    
    (c) Which variables contribute to the model? To see this, you can use 
    \texttt{coef()} for the output object. You obtain an object with 4 (= number 
    of groups) list elements, containing the estimated regression coefficients. 
    Thus, this is different from our approach to logistic regression with $K$ 
    groups in the course notes, where you would only obtain $K-1$ coefficient 
    vectors.
    
    ```{r, echo=TRUE}
    multi.glm.betas <- coef(mod.multi.glm) #  1.se
    multi.glm.beta.1 <- multi.glm.betas$`1`
    multi.glm.beta.2 <- multi.glm.betas$`2`
    multi.glm.beta.3 <- multi.glm.betas$`3`
    multi.glm.beta.4 <- multi.glm.betas$`4`
    
    print(paste("The coefficient vector for group 1 contains the variables",
                toString(multi.glm.beta.1@i[2:length(multi.glm.beta.1@i)])))
    print(paste("The coefficient vector for group 2 contains the variables",
                toString(multi.glm.beta.2@i[2:length(multi.glm.beta.2@i)])))
    print(paste("The coefficient vector for group 3 contains the variables",
                toString(multi.glm.beta.3@i[2:length(multi.glm.beta.3@i)])))
    print(paste("The coefficient vector for group 4 contains the variables",
                toString(multi.glm.beta.4@i[2:length(multi.glm.beta.4@i)])))
    ```
    
    (d) Select one of the variables from 2(c) which is relevant e.g. for the 
    first group, and plot this variable against the response (using the training 
    data). What you should see is that the values of the first group clearly 
    differ from those of the other groups.
    
    ```{r, echo=TRUE}
    # V123 vs y
    plot(Khan$xtrain[,123], Khan$ytrain,
         ylab="y",
         xlab="V123",
         main="Train Set")
    ```
    
    (e) Now use the trained model and predict the group membership of the test 
    data. Be careful, \texttt{predict()} yields predictions for each observation 
    to each class, and you need to select the appropriate class. Report the 
    confusion table and the misclassification error for the test data.
    
    ```{r, echo=TRUE}
    Khan.test.pred <- predict(mod.multi.glm, Khan$xtest)

    # iterate by row and column to determine predicted class for each obs
    Khan.test.pred.mod <- numeric(length(Khan$ytest))
    for(i in 1:nrow(Khan.test.pred)) {
      class <- 0
      max.prob <- -Inf
      for(j in 1:ncol(Khan.test.pred)) {
        if(Khan.test.pred[i, j,] > max.prob) {
          class <- j
          max.prob <- Khan.test.pred[i, j,]
        }
      }
      Khan.test.pred.mod[i] = class
    }
    
    # confusion table
    Khan.confusion <- confusionMatrix(data=as.factor(Khan.test.pred.mod),
                                           reference=as.factor(Khan$ytest))
    Khan.confusion$table
    ```

    The confusion table shows that the model perfectly predicted each class
    i.e. the misclassification rate for each class is zero.

</div></pre>