---
title: "Exercise 3"
author: "Kyzer Gerez"
date: "29.10.2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Advanced Methods for Regression and Classification
---

```{r, echo=TRUE}
load("building.RData")

# split into 2/3 train 1/3 test
smp_size <- floor(2/3 * nrow(df))

# set the seed to make partition reproducible
set.seed(1)
train_ind <- sample(seq_len(nrow(df)), size = smp_size)

train <- df[train_ind, ]
test <- df[-train_ind, ]
```

(1) \textit{Principal component regression (PCR)}:
    
    (a) Use the training data to fit a PCR model. PCR is implemented in the \texttt{library(pls)}
as the function \texttt{pcr()}, see help. Perform cross-validation using 10 segments
(see help of \texttt{pcr()}) and scale the data (\texttt{scale=TRUE}).
  
    ```{r, echo=TRUE, message=FALSE}
    library(pls)
    
    model.pcr <- pcr(y ~ .,
               data=train,
               validation="CV",
               segments=10,
               segment.type="random",
               scale=TRUE)
    
    summary(model.pcr)
    ```
    
    (b) Plot the obtained prediction errors from cross-validation, see lecture notes.
How many components seem to be optimal? What is the resulting RMSE?

    ```{r, echo=TRUE, warning=FALSE}
    # find optimal number of components based on one sigma
    cv_results <- RMSEP(model.pcr, estimate = "CV")
    comp_opt <- selectNcomp(model.pcr, method = "onesigma", plot = TRUE)
    paste("The optimal number of components is",
          comp_opt,
          "with an RMSE of",
          cv_results$val[1,,comp_opt])
    
    # zoomed in plot
    validationplot(model.pcr, val.type = "RMSEP", ncomp = 1:70, 
                   main = "Zoomed in Validation Plot")
    abline(v = comp_opt)
    ```

    Based on lecture discussions, the optimal number of components can be found
    by determining the minimum RMSE, then selecting the number of components below
    that which achieved the minimum RMSE while being within one standard error
    of this value.

    (c) Use the function \texttt{predplot()} to plot the measured \textit{y} values against the cross-validated \textit{y} values considering the optimal model.

    ```{r, echo=TRUE}
    predplot(model.pcr, comp_opt, line=TRUE)
    ```

    (d) Plot the predicted versus observed values for the test data, and compute the
RMSE. Compare with the values from last exercise.

    ```{r, echo=TRUE}
    library(Metrics)
    
    test_pcr_pred <- predict(model.pcr, test, comp_opt)
    test_pcr_rmse <- round(rmse(test$y, test_pcr_pred), 3)
    paste("Test RMSE for",
          comp_opt,
          "Comps:", 
          test_pcr_rmse)
    
    plot(test$y, test_pcr_pred)
    ```

    The forward selection model was the best performing model from the last exercise, 
    which achieved a test RMSE of 0.236. This model also performed better than the 
    optimal pcr model that resulted in a test RMSE of 0.357.
    
    The forward model used only 16 variables out of the 107 available. 
    PCR selects components that have the largest variance while
    variable selection looks directly at the relationship between a predictor
    and the response. For this dataset and train/test split, it happens that
    the choosing high variance components led to a model with less predictive
    power than simply directly looking at the relationship between predictors
    and the response.

    (e) The output object of \texttt{pcr()} contains the list elements \texttt{\$scores} and \texttt{\$loadings},
which are the matrices \textbf{Z} and \textbf{V} from the course notes. Visualize in scatterplots
the first two score vectors as well as the first two loadings vectors and try to
explain shortly what you see.

    ```{r, echo=TRUE}
    # Plot PC1 vs PC2 scores
    scores <- model.pcr$scores
    plot(scores[, 1], scores[, 2],
         xlab = "PC1 (Score 1)",
         ylab = "PC2 (Score 2)",
         main = "Scores Plot")
    abline(v=0)
    abline(h=0)
    
    # Plot PC1 vs PC2 loadings
    loadings <- model.pcr$loadings
    plot(loadings[, 1], loadings[, 2],
         xlab = "PC1 (Loading 1)",
         ylab = "PC2 (Loading 2)",
         main = "Loadings Plot")
    abline(v=0)
    abline(h=0)
    
    text(loadings[, 1], loadings[, 2], 
     labels = rownames(loadings),
     pos = 1, cex = 0.3, col = "darkred")
    ```

    First looking at the plot of scores, observations that cluster together are similar
    in the original high-dimensional space. The plot shows many clusters with a few
    points each. Also, the plot shows outliers such as the points along the lower
    left quadrant, which have large absolute values along both axes.
    
    The loadings plot shows how variables contribute to the principal components,
    with variables farther away from the origin having larger contributions.
    Additionally, variables that cluster together are highly correlated.
    The plot shows many variables that are highly correlated around (-0.12, 0.0).
    In general, variables that are lagged versions of another (e.g. Econ10.lag1,
    Econ10.lag2, etc.) are similar.

(2) \textit{$l_0$-norm regression}:

    Take a quick look at Equation (1) in https://arxiv.org/pdf/2202.04820. This $l_0$
    norm regularization has the effect that the size of the tuning parameter $\lambda$ determines
    the number of variables included in the model. Thus, this is a convenient way to do
    variable selection. We only need to identify an appropriate parameter $\lambda$.
    This choice can be made with the function \texttt{L0Learn.cvfit} of the package \texttt{L0Learn}.
    You need to provide \textbf{X} as matrix (\texttt{as.matrix()}), the response, and you can set a
    maximum size of the model (parameter \texttt{maxSuppSize}).

    (a) Use the training data to fit the model. Use \texttt{plot()} on the result object. How
can you interpret the plot? How can you select the optimal model?

    ```{r, echo=TRUE}
    library(L0Learn)
    
    L0.model <- L0Learn.cvfit(as.matrix(subset(train, select = -c(y))), train$y)
    plot(L0.model)
    ```

    The plot shows the mean cross-validation error model sizes ranging from
    0 to the \texttt{maxSuppSize}. Though \texttt{cvfit} does not create models 
    of every possible size between those two values. For each point, the mean
    error and standard deviation is shown.
    
    The optimal model can be selected by first looking for the one with the smallest
    mean error. Then, choosing the smallest possible model that is still within
    one standard error of the smallest mean error.

    (b) Which is the optimal tuning parameter $\lambda$. Extract the regression coefficients
for the optimal $\lambda$. Which variables are in the model? \textit{Hint}: See help of the
function.

    ```{r, echo=TRUE}
    # find minimum error index and its sd
    min_index <- which.min(unlist(L0.model$cvMeans[,]))
    min_RMSE <- L0.model$cvMeans[[1]][min_index]
    min_SD <- L0.model$cvSDs[[1]][min_index]
    
    # find index of optimal
    rmse_values <- L0.model$cvMeans[[1]][1:min_index] # trim RMSEs after min_index
    rmse_values <- rmse_values - min_RMSE - min_SD # find largest negative val
    opt_index <- 13 # by inspection of rmse_values
    
    # find optimal lambda
    lambda_opt <- L0.model$fit$lambda[[1]][opt_index]
    paste("The optimal tuning parameter is", lambda_opt)
    
    # extract coefs and vars from optimal model
    sparse_matrix <- coef(L0.model, lambda = lambda_opt)
    
    # convert sparse matrix to vector
    beta_vector <- as.vector(sparse_matrix)
    
    # get coefs and their indices
    nonzero_indices <- which(beta_vector != 0)
    variable_indices <- nonzero_indices[nonzero_indices > 1] - 1 # shift because of intercept
    l0_optimal_coefs <- beta_vector[nonzero_indices][-1] # remove intercept
    
    # Create df
    l0_optimal_coefs_df <- data.frame(
      variable = colnames(df[-1])[variable_indices],  # remove y
      coefficient = l0_optimal_coefs
      )
    l0_optimal_coefs_df
    ```

    (c) Plot the fitted values against the response for the training data and compute
the RMSE. Compare to PCR.

    ```{r, echo=TRUE}
    # l0
    train_l0_pred <- predict(L0.model, 
                          newx = as.matrix(subset(train, select = -c(y))), 
                          lambda = lambda_opt)
    train_l0_rmse <- round(rmse(train$y, as.vector(train_l0_pred)), 3)
    
    # PCR
    train_pcr_pred <- predict(model.pcr, train, comp_opt)
    train_pcr_rmse <- round(rmse(train$y, train_pcr_pred), 3)
    
    # l0 plotting
    plot(train$y, train_l0_pred,
         xlab = "Response",
         ylab = "Fitted",
         main = "l0 Fitted vs Response for Train"
         )
    abline(a = 0, b = 1)
    
    # output
    paste("PCR train RMSE:", train_pcr_rmse)
    paste("L0 train RMSE:", train_l0_rmse)
    
    ```

    Both models performed similarly on the train set, though the $l_0$ model was
    slightly better.

    (d) Plot the fitted values against the response for the test data and compute the
RMSE. Draw conclusions.
    
    ```{r, echo=TRUE}
    # l0
    test_l0_pred <- predict(L0.model, 
                      newx = as.matrix(subset(test, select = -c(y))), 
                      lambda = lambda_opt)
    test_l0_rmse <- round(rmse(test$y, as.vector(test_l0_pred)), 3)
    
    # l0 plotting
    plot(test$y, test_l0_pred,
         xlab = "Response",
         ylab = "Fitted",
         main = "l0 Fitted vs Response for Test"
         )
    abline(a = 0, b = 1)
    
    # PCR vs l0 comparison
    paste("PCR test RMSE:", test_pcr_rmse)
    paste("L0 test RMSE:", test_l0_rmse)
    
    ```

    Both models were again fairly similar in performance, with the $l_0$ model
    being slightly better. The forward selection model in the previous exercise
    was shown to be better than PCR, and $l_0$ functions similarly in that it
    explicitly selects predictors based on their relationship with the response.
    Since both the forward selection and $l_0$ models contained a small fraction
    of the predictors, the predictive power of PCR may have been watered down
    since the principal components incorporates all variables.
    These reasons could then explain why $l_0$ outperformed PCR.

    (e) Compare the regression coefficients (without intercept) from PCR and $l_0$-norm
regression in a plot and draw conclusions.

    ```{r, echo=TRUE}
    # load optimal pcr coefs into df
    pcr_optimal_coefs <- model.pcr$coefficients[,,"27 comps"]
    pcr_optimal_coefs_df <- data.frame(
      variable = names(pcr_optimal_coefs),
      coefficient = unname(pcr_optimal_coefs)
      )
    
    # combine dfs
    pcr_optimal_coefs_df$model <- "PCR"
    l0_optimal_coefs_df$model  <- "L0"
    coef_df <- rbind(pcr_optimal_coefs_df, l0_optimal_coefs_df)
    ```
    
    ```{r, echo=TRUE}
    library(ggplot2)
    
    ggplot(coef_df, aes(x = variable, y = coefficient, fill = model)) +
      geom_point(aes(color = model)) +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 3)) +
      labs(title = "Comparison of Regression Coefficients",
           x = "Variable",
           y = "Coefficient Estimate")
    ```

    In general, $l_0$ is sparse compared to PCR. This shows that only a few
    predictors are relevant in estimating the response. The PCR model has
    many variables where the coefficients were not driven down to zero compared
    to $l_0$. This indicates that the model may be capturing noise or using
    irrelevant predictors. There were a few variables where PCR estimated 
    especially large coefficients compared to $l_0$ such as \texttt{PhysFin1, PhyFin4, and
    PhysFin8}. These variables could have outsized contributions to the variance
    compared to the predictive power.

</div></pre>