---
title: "Exercise 5"
author: "Kyzer Gerez"
date: "12.11.2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Advanced Methods for Regression and Classification
---

Compute the following models by using the function \texttt{glmnet()} of the 
package \texttt{glmnet}, and by selecting the appropriate parameters.

```{r, echo=TRUE, message=FALSE}
library(robustHD)
library(glmnet)
library(Metrics)
set.seed(1)

# load dataset
data(nci60, package="robustHD")

# randomly select column from protein
protein_rand_ind <- sample(ncol(protein), 1)
protein_rand <- protein[, protein_rand_ind]

# randomly split into 3/4 train, 1/4 test
smp_size <- floor(3/4 * nrow(gene))
train_ind <- sample(seq_len(nrow(gene)), size = smp_size)

train_x <- as.matrix(gene[train_ind, ])
train_y <- as.matrix(protein_rand[train_ind])

test_x <- as.matrix(gene[-train_ind, ])
test_y <- as.matrix(protein_rand[-train_ind])
```

1. \textit{Ridge Regression}

```{r, echo=TRUE}
model.ridge <- glmnet(train_x, train_y,
                      alpha=0)

plot(model.ridge, sign.lambda = 1)
```

```{r, echo=TRUE}
cv.ridge <- cv.glmnet(train_x, train_y,
                      alpha=0)

plot(cv.ridge, sign.lambda=1)

lambda_max <- cv.ridge$lambda[1]
lambda_min <- tail(cv.ridge$lambda, n=1)
```
    
```{r, echo=TRUE}
# lambda_max
test_ridge_max_pred <- predict(cv.ridge, test_x, lambda_max)
test_ridge_max_rmse <- round(rmse(test_y, test_ridge_max_pred), 3)
paste("Test RMSE for ridge regression model with lambda =",
      round(lambda_max, 3),
      "is", 
      test_ridge_max_rmse)

plot(test_y, test_ridge_max_pred,
     main=paste("Actual vs Fitted for Test Using Ridge for lambda =", 
                round(lambda_max, 3)),
     xlab="Actual",
     ylab="Fitted")
abline(0, 1)

# lambda_min
test_ridge_min_pred <- predict(cv.ridge, test_x, lambda_min)
test_ridge_min_rmse <- round(rmse(test_y, test_ridge_min_pred), 3)
paste("Test RMSE for ridge regression model with lambda =",
      round(lambda_min, 3),
      "is", 
      test_ridge_min_rmse)

plot(test_y, test_ridge_min_pred,
     main=paste("Fitted vs Actual for Test Using Ridge for lambda =", 
                round(lambda_min, 3)),
     xlab="Actual",
     ylab="Fitted")
abline(0, 1)
```

The CV plot suggests that the optimal value for $\lambda$ is 745, which was the 
largest value calculated. However, when then trying to predict against the test
set, the model with this $\lambda$ value simply predicts the same value for all
observations, which is around the mean for `train_y`. This implies that
all coefficients were suppressed such that the model was effectively empty. Though
all predictors still had non-zero coefficients.

A better choice for $\lambda$ is 10, which was the minimum value calculated.
In this case, there are some predictors with relatively large coefficients.
This model performs better than the empty model with the maximum $\lambda$.

2. \textit{Lasso Regression}

```{r, echo=TRUE}
model.lasso <- glmnet(train_x, train_y,
                      alpha=1)

plot(model.lasso, sign.lambda = 1)
```

```{r, echo=TRUE}
cv.lasso <- cv.glmnet(train_x, train_y,
                      alpha=1)

plot(cv.lasso, sign.lambda=1)

# find lambda that is less than lambda.max but within 1 se
mse_threshold <- cv.lasso$cvm[1] + cv.lasso$cvsd[1]
for(i in 1:length(cv.lasso$cvm)) {
  if (cv.lasso$cvm[i] >= mse_threshold) {
    opt_index_lasso <- i
    break
  }
}
opt_lambda_lasso <- cv.lasso$lambda[opt_index_lasso]

print(paste("The optimal model contains", cv.lasso$glmnet.fit$df[opt_index_lasso], 
            "non-zero coefficients"))
```

```{r, echo=TRUE}
# lambda_max
test_lasso_pred <- predict(cv.lasso, test_x, opt_lambda_lasso)
test_lasso_rmse <- round(rmse(test_y, test_lasso_pred), 3)
paste("Test RMSE for lasso regression model with lambda =",
      round(opt_lambda_lasso, 3),
      "is", 
      test_lasso_rmse)

plot(test_y, test_lasso_pred,
     main=paste("Fitted vs Actual for Test Using Lasso for lambda =", 
                round(opt_lambda_lasso, 3)),
     xlab="Actual",
     ylab="Fitted")
abline(0, 1)
```

A modified version of the 1se rule was used to find the optimal lambda for the 
lasso regression model. The difference is that the smallest lambda is selected
instead of the largest with respect to the standard error of the minimum MSE.
The optimal model contained `r cv.lasso$glmnet.fit$df[opt_index_lasso]` non-zero 
coefficients out of the original 22,283 predictors.

3. \textit{Elastic Net Regression}

```{r, echo=TRUE}
# generate models with alpha in increments of 0.05 from [0,1]
# and collect min cvm for all models
elastic.net.models <- list()
min_cvms <- numeric(0)
for (i in 1:19) {
  name <- paste0("alpha", i/20)
  
  elastic.net.models[[name]] <-
    cv.glmnet(train_x, train_y, type.measure="mse", alpha=i/20)
  
  # extract min cvm
  min_index <- elastic.net.models[[name]]$index[1]
  min_cvms[[name]] <- elastic.net.models[[name]]$cvm[min_index]
}

# find model with smallest min cvm
opt.alpha <- names(min_cvms)[min_cvms == min(min_cvms)]
opt.model.elastic <- elastic.net.models[[opt.alpha]]

# determine mse threshold using 1se rule
min_index <- opt.model.elastic$index[1]
mse_threshold <- opt.model.elastic$cvm[min_index] + opt.model.elastic$cvsd[min_index]

# iterate and find opt_lambda
for(i in 1:length(opt.model.elastic$cvm)) {
  if (opt.model.elastic$cvm[i] >= mse_threshold) {
    opt_index_elastic <- i
    break
  }
}
opt_lambda_elastic <- opt.model.elastic$lambda[opt_index_elastic]

print(paste("The optimal model is", 
            opt.alpha, 
            "with lambda =", 
            round(opt_lambda_elastic, 3)))

plot(opt.model.elastic, sign.lambda=1)

print(paste("The optimal model contains", opt.model.elastic$nzero[opt_index_elastic], 
            "non-zero coefficients"))
```

```{r, echo=TRUE}
test_elastic <- predict(opt.model.elastic, test_x, opt_lambda_elastic)
test_elastic_rmse <- round(rmse(test_y, test_elastic), 3)
paste("Test RMSE for elastic model with",
      opt.alpha,
      "and lambda =",
      round(opt_lambda_lasso, 3),
      "is", 
      test_elastic_rmse)

plot(test_y, test_elastic,
     main=paste("Fitted vs Actual for Test for", 
                opt.alpha,
                "and lambda =", 
                round(opt_lambda_elastic, 3)),
     xlab="Actual",
     ylab="Fitted")
abline(0, 1)
```

4. \textit{Adaptive Lasso Regression}

```{r, echo=TRUE}
# get inverse of the absolute Ridge coefficients
coef.ridge <- coef(cv.ridge, lambda_min)

# calculate adaptive lasso regression model
alasso <- glmnet(train_x, train_y, penalty.factor = 1 / abs(coef.ridge[-1]))

plot(alasso, xvar="lambda", sign.lambda=1)
```
```{r, echo=TRUE}
# run cv to find optimal tuning parameter
alasso.cv <- cv.glmnet(train_x, train_y, penalty.factor = 1 / abs(coef.ridge[-1]))

plot(alasso.cv, sign.lambda=1)
```
```{r, echo=TRUE}
pred.alasso <- predict(alasso.cv, newx=test_x, s="lambda.1se")
pred.alasso.rmse <- round(rmse(test_y, pred.alasso), 3)
paste("Test RMSE for adaptive Lasso model:",
      pred.alasso.rmse)

plot(test_y, pred.alasso,
     main="Fitted vs Actual for Test for Adaptive Lasso Model",
     xlab="Actual",
     ylab="Fitted")
abline(0, 1)
```

5. Which of the models was most useful for the prediction task, which was not useful?

The elastic net regression model with `r opt.alpha` and $\lambda=$ 
`r opt_lambda_elastic` achieved the lowest RMSE in predicting the test set. 
The optimal model was determined by first finding the $\alpha$ value that 
resulted in a model with the lowest cvm. Then, a modified version of the 1 se 
rule was used where instead the smallest possible lambda is selected. If the 1 
se rule was followed, a model with only one coefficient would've been selected 
instead.

The Ridge and Lasso regression models also performed reasonably well, if the
modified 1 se rule was followed.

The adaptive Lasso regression model was not useful since it performed worse than
simply predicting the average of the train set.

6. The practitioner not only wants good predictions, but also interpretability of the
model. Which of the models was most useful for this task?

The Lasso regression model is by far the most interpretable in that it only
contains `r cv.lasso$glmnet.fit$df[opt_index_lasso]` non-zero coefficients.
The elastic model performed better, but it contained 
`r opt.model.elastic$nzero[opt_index_elastic]` 
non-zero coefficients in contrast. 

The Ridge regression model is not useful at all since coefficients are not driven
down to zero.

</div></pre>