---
title: "Exercise 2"
author: "Kyzer Gerez"
date: "23.10.2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Advanced Methods for Regression and Classification
---

```{r, echo=TRUE}
load("building.RData")

# split into 2/3 train 1/3 test
smp_size <- floor(2/3 * nrow(df))

# set the seed to make partition reproducible
set.seed(1)
train_ind <- sample(seq_len(nrow(df)), size = smp_size)

train <- df[train_ind, ]
test <- df[-train_ind, ]
```

(1) Compute the full model with lm() for the training data, and predict the response
values for the test data.

    ```{r, echo=TRUE}
    full_model <- lm(y ~ ., train)
    test_pred <- predict(full_model, test)
    ```
    
    (a) Use plot() for the result object and briefly discuss the findings.
  
    ```{r, echo=TRUE}
    plot(full_model)
    ```
  
    The residuals vs fitted plot indicates some nonlinearity when looking at
    and above the fitted value 7. The Q-Q plot suggests that the residuals are
    normally distributed, with the only outlier being observation 29, which was also 
    noted in the previous plot. The scale-location plot shows that the assumption
    of equal variance may not hold, with a wider spread along the middle of the
    x-axis range. No datapoints standout in the residuals vs leverage plot. This
    indicates that datapoint 29 was not influential despite being notable in the
    other plots.
    
    (b) Look at the estimated regression coefficients. Depending on your training/test
split these might include NA’s. What could be the reason for that?

    ```{r, echo=TRUE}
    summary(full_model)
    ```

        35 predictors don't have defined coefficients because each of them have an exact
        linear relationship with at least one other predictor.

    (c) Compute the RMSE (=square-root of the MSE) for training and test data, and
discuss the findings.

    ```{r, echo=TRUE}
    library('Metrics')
    
    train_pred <- predict(full_model, train)
    
    precision <- 3
    train_RMSE <- round(rmse(train$y, train_pred), precision)
    test_RMSE <- round(rmse(test$y, test_pred), precision)
                      
    print(paste('Train RMSE:', train_RMSE))
    print(paste('Test RMSE:', test_RMSE))
    ```

        The test RMSE is significantly larger than the train RMSE, which suggests that
        the full model has an overfitting problem/high variance.

    (d) Plot the observed response values versus the fitted/predicted values for training
and test data, and discuss the findings.

    ```{r, echo=TRUE}
    library(ggplot2)
    
    # train
    ggplot(data.frame(actual = train$y, predicted = train_pred),
      aes(x = actual, y = predicted)) +
      ylim(-10, 10) +
      geom_point() +
      geom_abline(color = "red") +
      labs(title = "Training Set: Predicted vs Actual y",
        x = "Actual y", y = "Predicted y")
    
    # test
    ggplot(data.frame(actual = test$y, predicted = test_pred),
      aes(x = actual, y = predicted)) +
      ylim(-10, 10) +
      geom_point() +
      geom_abline(color = "red") +
      labs(title = "Test Set: Predicted vs Actual y",
        x = "Actual y", y = "Predicted y")
    ```

        The training set plot shows that the predicted y is generally close to
        the actual y. At the higher end of actual y, the predicted y does tend
        to overestimate though.
        
        The test set shows much larger residuals across the scale, with one
        data point having a predicted y that is far below the actual y.

(2) Perform variable selection based on stepwise regression, using the function step(),
see help file and course notes. Perform both, forward selection (start from the empty
model) and backward selection (start from the full model). Compare the resulting
models with the RMSE, and with plots of response versus predicted values.

    ```{r, echo=TRUE}
    forward_selection <- step(lm(y ~ 1, data = df), 
                              direction = "forward", 
                              scope = formula(full_model),
                              trace=0)
    backward_selection <- step(full_model, 
                               direction = "backward", 
                               trace=0)
    ```
    
    ```{r, echo=TRUE}
    # train - predict
    forward_train_pred <- predict(forward_selection, train)
    backward_train_pred <- predict(backward_selection, train)
    
    # train - RMSE
    forward_train_RMSE <- round(rmse(train$y, forward_train_pred), precision)
    backward_train_RMSE <- round(rmse(train$y, backward_train_pred), precision)
    
    # test - predict
    forward_test_pred <- predict(forward_selection, test)
    backward_test_pred <- predict(backward_selection, test)
    
    # test - RMSE
    forward_test_RMSE <- round(rmse(test$y, forward_test_pred), precision)
    backward_test_RMSE <- round(rmse(test$y, backward_test_pred), precision)
    
    # print
    print(paste('Forward train RMSE:', forward_train_RMSE))
    print(paste('Backward train RMSE:', backward_train_RMSE))
    
    print(paste('Forward test RMSE:', forward_test_RMSE))
    print(paste('Backward test RMSE:', backward_test_RMSE))
    
    # plotting - test only
    ggplot(data.frame(actual = test$y, predicted = forward_test_pred),
    aes(x = actual, y = predicted)) +
    ylim(-10, 10) +
    geom_point() +
    geom_abline(color = "red") +
    labs(title = "Forward Selection Test Set: Predicted vs Actual y",
      x = "Actual y", y = "Predicted y")
    
    ggplot(data.frame(actual = test$y, predicted = backward_test_pred),
    aes(x = actual, y = predicted)) +
    ylim(-10, 10) +
    geom_point() +
    geom_abline(color = "red") +
    labs(title = "Backward Selection Test Set: Predicted vs Actual y",
      x = "Actual y", y = "Predicted y")
    ```

(3) Which of the models from (1) and (2) would be the preferred one, and why? Would
a model comparison based on ANOVA be useful?

    Looking at the RMSE and the prediced vs actual y plots, the forward 
    selection model greatly outperforms the full model and backward selection model.
    A model comparison based on ANOVA seems to be unnecssary since the simpler
    model based on forward selection is much better. ANOVA is useful when
    the more complex model outperforms and a test is needed to determine
    if this outperformance justifies the increased complexity.

(4) Use the function cvFit() from the package cvTools to compare the models from (1) 
and (2), separately for training and test set observations (thus 6 results). A
look at the help file shows how you can use the function for model evaluation based
on cross-validation (CV). Set the parameters cost=rmspe for RMSE calculations,
and perform 5-fold CV with 100 replications. Compare all 6 error distributions
(available in $reps) by parallel boxplots. What do you conclude?

    ```{r, echo=TRUE, warning=FALSE}
    library(cvTools)
    
    # --- train ---
    full_cv_RMSE_train <- cvFit(full_model, data=train, y=train$y, 
                                cost=rmspe, K=5, R=100, seed=1)
    forward_cv_RMSE_train <- cvFit(forward_selection, data=train, y=train$y, 
                            cost=rmspe, K=5, R=100, seed=1)
    backward_cv_RMSE_train <- cvFit(backward_selection, data=train, y=train$y, 
                        cost=rmspe, K=5, R=100, seed=1)
        
    # --- test ---
    full_cv_RMSE_test <- cvFit(full_model, data=test, y=test$y, 
                                cost=rmspe, K=5, R=100, seed=1)
    forward_cv_RMSE_test <- cvFit(forward_selection, data=test, y=test$y, 
                            cost=rmspe, K=5, R=100, seed=1)
    backward_cv_RMSE_test <- cvFit(backward_selection, data=test, y=test$y, 
                        cost=rmspe, K=5, R=100, seed=1)
    ```

    ```{r, echo=TRUE}
    # create dfs for plotting
    train_errors_df <- class(data.frame)
    train_errors_df$rmse <- data.frame(
      Error = c(full_cv_RMSE_train$reps, 
                forward_cv_RMSE_train$reps, 
                backward_cv_RMSE_train$reps),
      Model = factor(rep(c("Full", "Forward", "Backward"),
                         times = c(length(full_cv_RMSE_train$reps), 
                                   length(forward_cv_RMSE_train$reps), 
                                   length(backward_cv_RMSE_train$reps)))))
    
    test_errors_df <- class(data.frame)
    test_errors_df$rmse <- data.frame(
      Error = c(full_cv_RMSE_test$reps, 
                forward_cv_RMSE_test$reps, 
                backward_cv_RMSE_test$reps),
      Model = factor(rep(c("Full", "Forward", "Backward"),
                         times = c(length(full_cv_RMSE_test$reps), 
                                   length(forward_cv_RMSE_test$reps), 
                                   length(backward_cv_RMSE_test$reps)))))
    ```
    
    ```{r, echo=TRUE}
    library(ggplot2)

    ggplot(train_errors_df$rmse, aes(x = Model, y = Error, fill = Model)) +
      geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2) +
      scale_y_continuous(trans='log10') +
      labs(
        title = "CV Error Distributions by Model for Train",
        y = "CV Error (log)",
        x = "Model"
      ) +
      theme_minimal() +
      theme(legend.position = "none")
    
    ggplot(test_errors_df$rmse, aes(x = Model, y = Error, fill = Model)) +
      scale_y_continuous(trans='log10') +
      geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2) +
      labs(
        title = "CV Error Distributions by Model for Test",
        y = "CV Error (log)",
        x = "Model"
      ) +
      theme_minimal() +
      theme(legend.position = "none")
    ```

    The full model is far more unstable due to significant outliers compared
    to the variable selection models. The forward model is much
    more stable over all replications and across different folds, and it has 
    the lowest median error for both the train and test sets. The backward model 
    is an improvement over the full model but still suffers from overfitting and 
    sensitivity to outliers.

(5) Do the same as in (4), but use cost=rtmspe. What are we doing? Which model is
the preferred one?

    ```{r, echo=TRUE, warning=FALSE}
    # --- train ---
    full_cv_RTMSE_train <- cvFit(full_model, data=train, y=train$y, 
                                cost=rtmspe, K=5, R=100, seed=1)
    forward_cv_RTMSE_train <- cvFit(forward_selection, data=train, y=train$y, 
                            cost=rtmspe, K=5, R=100, seed=1)
    backward_cv_RTMSE_train <- cvFit(backward_selection, data=train, y=train$y, 
                        cost=rtmspe, K=5, R=100, seed=1)
        
    # --- test ---
    full_cv_RTMSE_test <- cvFit(full_model, data=test, y=test$y, 
                                cost=rtmspe, K=5, R=100, seed=1)
    forward_cv_RTMSE_test <- cvFit(forward_selection, data=test, y=test$y, 
                            cost=rtmspe, K=5, R=100, seed=1)
    backward_cv_RTMSE_test <- cvFit(backward_selection, data=test, y=test$y, 
                        cost=rtmspe, K=5, R=100, seed=1)
    ```

    ```{r, echo=TRUE}
    # create dfs for plotting
    train_errors_df$rtmse <- data.frame(
      Error = c(full_cv_RTMSE_train$reps, 
                forward_cv_RTMSE_train$reps, 
                backward_cv_RTMSE_train$reps),
      Model = factor(rep(c("Full", "Forward", "Backward"),
                         times = c(length(full_cv_RTMSE_train$reps), 
                                   length(forward_cv_RTMSE_train$reps), 
                                   length(backward_cv_RTMSE_train$reps)))))
    
      test_errors_df$rtmse <- data.frame(
      Error = c(full_cv_RTMSE_test$reps, 
                forward_cv_RTMSE_test$reps, 
                backward_cv_RTMSE_test$reps),
      Model = factor(rep(c("Full", "Forward", "Backward"),
                         times = c(length(full_cv_RTMSE_test$reps), 
                                   length(forward_cv_RTMSE_test$reps), 
                                   length(backward_cv_RTMSE_test$reps)))))
    ```

    ```{r, echo=TRUE}
    ggplot(train_errors_df$rtmse, aes(x = Model, y = Error, fill = Model)) +
      geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2) +
      labs(
        title = "CV Error Distributions by Model for Train",
        y = "CV Error",
        x = "Model"
      ) +
      theme_minimal() +
      theme(legend.position = "none")
    
    ggplot(test_errors_df$rtmse, aes(x = Model, y = Error, fill = Model)) +
      geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2) +
      labs(
        title = "CV Error Distributions by Model for Test",
        y = "CV Error",
        x = "Model"
      ) +
      theme_minimal() +
      theme(legend.position = "none")
    ```

    RTMSPE is a cost function that trims the largest squared differences of
    the observed and fitted values (25% by default). This means that outliers
    are excluded. Because of this, the backward and full models don't appear
    as unstable compared to using RMSPE. Still, the forward model is superior
    to the other two.

</div></pre>