---
title: "Exercise 10"
author: "Kyzer Gerez"
date: "07.01.2026"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Advanced Methods for Regression and Classification
---

Load the data \texttt{Diabetes} from the package \texttt{ROCit}. Our goal is to 
find a classification model for \texttt{diabetes}, based on the variable 
\texttt{dtest} – see help file. Convert this information to a binary variable. 
To simplify the task, use as predictor variables: 
\texttt{chol, stab.glu, hdl, bmi, age,} and \texttt{gender}. Afterwards, delete 
observations with missings, using \texttt{na.omit()}.

```{r echo=TRUE, message=FALSE}
library(ROCit); data(Diabetes)
library(mgcv)
library(caret)

# convert dtest to binary variable
Diabetes$dtest <- ifelse(Diabetes$dtest == "+", TRUE, FALSE)

# select only required predictors
vars <- c("chol", "stab.glu", "hdl", "bmi", "age", "gender", "dtest")
Diabetes <- Diabetes[vars]

# delete obs with NAs
Diabetes <- na.omit(Diabetes)
```

Select randomly a training set of about 3/4 of the observations from each class, 
build the classification model, predict the group membership for the (remaining) 
test data and, as the final result, show the resulting confusion table and 
compute the misclassification rate.

```{r, echo=TRUE}
set.seed(12427543)

# create 3/4 train and test sets
samp_size <- floor(3/4*nrow(Diabetes))
train_indx <- sample(1:nrow(Diabetes), samp_size)
train <- Diabetes[train_indx,]
test <- Diabetes[-train_indx,]
```

(1) Fit a logistic regression model, see Exercise 7, and report the results.
    
    ```{r, echo=TRUE}
    # train logistic regression model
    mod.log.reg <- glm(dtest ~., train, family="binomial")
    
    # predict on test
    log.reg.pred <- predict(mod.log.reg, test)
    
    # convert from linear to binary scale
    log.reg.pred <- ifelse(log.reg.pred<=0, FALSE, TRUE)
    
    # build confusion matrix
    log.reg.confusion <- confusionMatrix(data=as.factor(log.reg.pred), 
                                         reference=as.factor(test$dtest))
    log.reg.confusion$table
    
    # compute misclassification rate
    log.reg.misclass <- (log.reg.confusion$table[2,1] + # FP
      log.reg.confusion$table[1,2]) / # FN
      (nrow(test))
    print(paste("The misclassification rate for logistic regression is", 
                round(log.reg.misclass, 3)))
    ```

(2) Use GAMs, implemented in the function \texttt{gam()} of the package 
\texttt{library(mgcv)}, with the argument \texttt{family="binomial"}.

    (a) The smooth functions in GAMs can be defined for every variable by 
    \texttt{s(variable)}, see also course notes. It might not make sense to use 
    smooth functions for all variables, for sure not for factor variables. Now 
    compute the GAM based on your chosen “formula”.
    
    ```{r, echo=TRUE}
    # train GAM
    gam.object <- gam(dtest~s(chol)+s(stab.glu)+s(hdl)+s(bmi)+s(age)+gender,
                      data=train,
                      family="binomial")
    ```
    
    (b) Which variables are significant in the model? How complex are the smooth 
    functions (in terms of the effective degrees of freedom)?
    
    ```{r, echo=TRUE}
    summary(gam.object)
    ```
    
    The variables \texttt{chol} and \texttt{stab.glu} are significant. The 
    smooth functions for the variables \texttt{stab.glu}, \texttt{bmi}, and 
    \texttt{age} are rather complex, having about 7-8 effective degrees of 
    freedom. In contrast, \texttt{hdl} only has 1 effective degree of freedom,
    so linear.
    
    (c) Plot the explanatory variables against their smoothed values as they are 
    used in the model. You can simply use:
    \texttt{plot(gam.object,page=1,shade=TRUE,shade.col="yellow")} How can you 
    interpret these plots?
    
    ```{r, echo=TRUE}
    plot(gam.object,page=1,shade=TRUE,shade.col="yellow")
    ```
    
    The plots show the estimates for each smooth term. The yellow band indicates
    the 95% confidence interval for the smooth estimate. The ticks on the x-axis
    represent each datapoint for that predictor. Sparse regions generally show
    larger CIs since the extrapolation risk is greater. The y-axis shows the 
    effect of the predictor on the response.
    
    (d) Report the confusion table and the misclassification error for the test 
    set. How does this compare to (1)?
    
    ```{r, echo=TRUE}
    # predict on test
    gam.pred <- predict(gam.object, test)
    
    # convert from linear to binary scale
    gam.pred <- ifelse(gam.pred<=0, FALSE, TRUE)
    
    # build confusion matrix
    gam.confusion <- confusionMatrix(data=as.factor(gam.pred), 
                                         reference=as.factor(test$dtest))
    gam.confusion$table
    
    # compute misclassification rate
    gam.misclass <- (gam.confusion$table[2,1] + # FP
      gam.confusion$table[1,2]) / # FN
      (nrow(test))
    print(paste("The misclassification rate for GAM is", 
                round(gam.misclass, 3)))
    ```
    GAM performed significantly worse than logistic regression due to having a 
    higher false positive rate.
    
    (e) The complexity of the smooth functions could be controlled by the 
    parameter \texttt{k} within \texttt{s()}, which allows to set an upper bound 
    for the effective degrees of freedom. If this makes sense, try it out and 
    check if the results in (2)(b) and (d) change.
    
    ```{r, echo=TRUE}
    # train GAM
    gam.object.mod <- gam(dtest~s(chol)+s(stab.glu,k=3)+s(hdl)+s(bmi)+s(age)+gender,
                      data=train,
                      family="binomial")
    
    # show model
    summary(gam.object.mod)
    
    # predict on test
    gam.mod.pred <- predict(gam.object.mod, test)
    
    # convert from linear to binary scale
    gam.mod.pred <- ifelse(gam.mod.pred<=0, FALSE, TRUE)
    
    # build confusion matrix
    gam.mod.confusion <- confusionMatrix(data=as.factor(gam.mod.pred), 
                                         reference=as.factor(test$dtest))
    gam.mod.confusion$table
    
    # compute misclassification rate
    gam.mod.misclass <- (gam.mod.confusion$table[2,1] + # FP
      gam.mod.confusion$table[1,2]) / # FN
      (nrow(test))
    print(paste("The misclassification rate for GAM with set EDoF is", 
                round(gam.mod.misclass, 3)))
    ```

    I tested different values of \texttt{k} for the variables \texttt{chol} and
    \texttt{stab.glu} since they were signifcant in the original GAM. I found
    that setting \texttt{s(stab.glu,k=3)} yielded a much better model with a
    misclassification rate of 9.4%, though this is still higher than the logistic
    regression model. Worth noting is that the modified model now lists 
    \texttt{stab.glu} as very significant.

</div></pre>