---
title: "Exercise 4"
author: "Kyzer Gerez"
date: "05.11.2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Advanced Methods for Regression and Classification
---

(1) We have realized that the data set is in fact a multivariate time series data set.
Argue why the training/test split we have used so far is not very appropriate for
such data, and use a more appropriate split for the following tasks.

    With random sampling, the model could be trained on future observations that are
then used to predict past observations. This is not appropriate since the model
will rely on data that is not yet available to make predictions. The proper
approach would be to split the data chronologically into train/test sets based
on the variables \texttt{COMPLETION.YEAR} and \texttt{COMPLETION.QUARTER}.

```{r, echo=TRUE}
load("building.RData")

# sort by COMPLETION YEAR and QUARTER
df <- sort_by(df, list(df$COMPLETION.YEAR, df$COMPLETION.QUARTER))

# split chronologically into 2/3 train 1/3 test
train_size <- floor(2/3 * nrow(df))

train <- df[1:train_size, ]
test <- df[(train_size+1):nrow(df), ]
```

(2) \textit{Partial Least Squares regression (PLS)}:
    
    (a) Use the training data to fit a PLS model. PLS is implemented in the \texttt{library(pls)} as the function \texttt{plsr()}, see help. Perform (useful?) cross-validation using 10 segments (see help of \texttt{plsr()}) and scale the data (\texttt{scale=TRUE}).
  
    ```{r, echo=TRUE, message=FALSE}
    library(pls)
    
    model.pls <- plsr(y ~ .,
               data=train,
               validation="CV",
               segments=10,
            #   segment.type="consecutive",
               scale=TRUE)
    
    summary(model.pls)
    ```
    
    Cross-validation is probably not appropriate since most of the splits
    will have test data that is dated prior to at least some training data.
    
    (b) Plot the obtained prediction errors from cross-validation, see lecture notes.
How many components seem to be optimal? What is the resulting RMSE?

    ```{r, echo=TRUE, warning=FALSE}
    # find optimal number of components based on one sigma
    cv_results <- RMSEP(model.pls, estimate = "CV")
    comp_opt <- selectNcomp(model.pls, method = "onesigma", plot = TRUE)
    paste("The optimal number of components is",
          comp_opt,
          "with an RMSE of",
          round(cv_results$val[1,,comp_opt], 3))
    
    # zoomed in plot
    validationplot(model.pls, val.type = "RMSEP", ncomp = 1:55, 
                   main = "Zoomed in Validation Plot")
    abline(v = comp_opt)
    ```

    Based on lecture discussions, the optimal number of components can be found
    by determining the minimum RMSE, then selecting the number of components below
    that which achieved the minimum RMSE while being within one standard error
    of this value.

    (c) Use the function \texttt{predplot()} to plot the measured \textit{y} values against the cross-validated \textit{y} values considering the optimal model.

    ```{r, echo=TRUE}
    predplot(model.pls, comp_opt, line=TRUE)
    ```

    (d) Plot the predicted versus observed values for the test data, and compute the
RMSE. Compare with the values from last exercise.

    ```{r, echo=TRUE}
    library(Metrics)
    
    test_pls_pred <- predict(model.pls, test, comp_opt)
    test_pls_rmse <- round(rmse(test$y, test_pls_pred), 3)
    paste("Test RMSE for",
          comp_opt,
          "Comps:", 
          test_pls_rmse)
    
    plot(test$y, test_pls_pred,
         main="Predicted vs Observed for Test Set",
         xlab="Observed",
         ylab="Predicted")
    abline(a=0, b=1)
    ```
    
    From the last exercise, the optimal PCR model based on the one standard 
    error rule achieved a test RMSE of 0.357. This is better than the optimal PLS
    model, which resulted in a test RMSE of 0.479. The difference in performance
    could be due to the train/test split. The PCR model was created based on a
    random split. It could be that there is some temporal trend that the PCR model
    took advantage of, which then led to a model that performs better than it should.

(3) \textit{Ridge Regression}:

    (a) Use the function \texttt{glmnet()} from the \texttt{library(glmnet)}
    with the parameter \texttt{alpha=0} and apply it to the training data – 
    see also course notes. Plot the result object. How can you interpret the plot? Which
    default parameters are used for \texttt{lambda}? What is the meaning of the parameter
    \texttt{alpha}?

    ```{r, echo=TRUE}
    library(glmnet)
    
    model.ridge <- glmnet(as.matrix(train[-c(1)]), train$y,
                          alpha=0)
    
    plot(model.ridge, sign.lambda = 1)
    
    print(model.ridge)
    ```

    Each curve in the plot represents the estimated coefficient for a variable
    at various values of \texttt{lambda}. As the penalty term increases, coefficients
    are drived down to zero. Printing the fitted model will show
    the number of non-zero coefficients at every \texttt{lambda} values. In this
    case, all variables have non-zero coefficients across all \texttt{lambda} 
    values.
    
    The parameter \texttt{lambda} is set to \texttt{NULL} by default. This means
    that the machine then computes its own sequence of \texttt{lambda} values
    based on the parameters \texttt{nlambda} and \texttt{lambda.min.ratio}.
    By default \texttt{nlambda=100}, so 100 values are generated. Since there
    are more observations than variables, \texttt{lambda.min.ratio} defaults to
    \texttt{1e-04}. This value is multiplied with \texttt{lambda.max} to get the
    smallest value for \texttt{lambda} in the sequence. \texttt{lambda.max}
    represents the smallest value for which all coefficients are zero.

    The parameter \texttt{alpha} is the elasticnet mixing parameter, which is 
    used to set the penalty. \texttt{alpha=0} is the ridge penalty, and 
    \texttt{alpha=1} is the lasso penalty.

    (b) Use the function \texttt{cv.glmnet()} and apply it to the training data. 
    Visualize and interpret the results. How do you obtain the optimal tuning 
    parameter and the regression coefficients?

    ```{r, echo=TRUE}
    cv.ridge <- cv.glmnet(as.matrix(train[-c(1)]), train$y,
                          alpha=0)
    
    plot(cv.ridge, sign.lambda=1)
    
    lambda_opt <- tail(cv.ridge$lambda, n=1)
    paste("The optimal tuning parameter is", lambda_opt)
    
    print("The regression coefficients for the optimal tuning parameter are:")
    coef(cv.ridge, s="lambda.min")
    ```

    For this case, the smallest \texttt{lambda} value is the optimal tuning parameter since
    it has the smallest MSE. More generally, the one standard error rule could
    be applied to find the optimal \texttt{lambda}. The regression coefficients
    can be found by using \texttt{coef} with \texttt{p} set to the optimal
    \texttt{lambda}.

    (c) Use the optimal model to predict the response for the test data. Compare 
    the predictions with the reported values graphically. Compare the RMSE results
    with those from the previous exercises.

    ```{r, echo=TRUE}
    test_ridge_pred <- predict(cv.ridge, as.matrix(test[-c(1)]), lambda_opt)
    test_ridge_rmse <- round(rmse(test$y, test_ridge_pred), 3)
    paste("Test RMSE for ridge regression model with lambda =",
          round(lambda_opt, 3),
          "is", 
          test_ridge_rmse)
    
    plot(test$y, test_ridge_pred,
         main="Actual vs Fitted for Test Using Ridge",
         xlab="Actual",
         ylab="Fitted")
    abline(0, 1)
    ```

    The plot is similar to the one for pls where there is some curvature 
    for the larger \texttt{y} values. The RMSE for the ridge model is fairly
    close to that for the PLS model but slightly worse. This then also means that
    the ridge model performed worse than the models from previous exercises
    that used random train and test splits.

    (d) Going back to 3.(a): At least my plot shows that the curves are not strictly
    increasing, but some decrease again after an increase. This seems counter-
    intuitive. What could be the explanation?

    It could be the case that some variables get driven down to zero, which
    leads to another variable suddenly becoming more important and estimated 
    a larger coefficient. The estimated coefficient then gets driven down again with
    larger \texttt{lambda}. This can happen when multicollinearity is present
    among the predictors, such as the lagged predictors in this dataset.

</div></pre>