---
title: "Exercise 8"
author: "Kyzer Gerez"
date: "03.12.2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Advanced Methods for Regression and Classification
---

```{r, echo=TRUE}
library(datasets)
data("cars")

# split into 2/3 train, 1/3 test
smp_size <- floor(2/3 * nrow(cars))

# set the seed to make partition reproducible
set.seed(1)
train_ind <- sample(seq_len(nrow(cars)), size = smp_size)

train <- cars[train_ind, ]
test <- cars[-train_ind, ]

# sort train and test sets by speed
train <- train[order(train$speed),]
test <- test[order(test$speed),]
```

1. Construct the basis functions from the training set. Visualize the basis 
functions.
    
    ```{r, echo=TRUE}
    lecturespl <- function(x, nknots=2, M=4) {
      # nknots ... number of knots -> placed at regular quantiles
      # M ... M-1 is the degree of the polynomial
      n <- length(x)
      # X will not get an intercept column
      X <- matrix(NA,nrow=n,ncol=(M-1)+nknots)
      for (i in 1:(M-1)){ 
        X[,i] <- x^i 
        }
      # now the basis functions for the constraints:
      quant <- seq(0,1,1/(nknots+1))[c(2:(nknots+1))]
      qu <- quantile(x,quant)
      for (i in M:(M+nknots-1)) {
        X[,i] <- ifelse(x-qu[i-M+1]<0,0,(x-qu[i-M+1])^(M-1))
      }
      list(X=X,quantiles=quant,xquantiles=qu)
      }
    
    # added so that basis functions can be evaluated on new x
    make_basis <- function(x, nknots=2, M=4) {
      out <- lecturespl(x, nknots, M)
      X <- out$X
      colnames(X) <- paste0("b", seq_len(ncol(X)))
      list(X = X, info = out)
      }
    
    plotspl <- function(splobj,...) {
      matplot(splobj$x,splobj$X,type="l",lty=1,
              xlab="x",ylab="h(x)",...)
      abline(v=splobj$xquantiles,lty=3,col=gray(0.5))
      }
    
    # create basis functions
    spl <- make_basis(train$speed)
    
    #spl <- lecturespl(train$speed)
    spl$x <- train$speed
    
    # visualize basis functions
    plotspl(spl)
    ```
    
2. Use these basis functions within \texttt{lm()} to predict the training data 
of the response. Plot the training data (dist versus speed) and visualize the 
predictions.

    ```{r, echo=TRUE}
    df_train <- data.frame(y = train$dist, spl$X)
    lm1 <- lm(y ~ ., data = df_train)
    
    train_pred <- predict(lm1, newdata=df_train)
    plot(train$speed, train$dist, col="blue", type="p",
         main="Training Data with Predictions",
         xlab="Distance",
         ylab="Speed")
    lines(train$speed, train_pred)
    ```
    
3. Use the model to predict the response from the test set observations. 
Visualize training and test set in one plot, using different colors, and present 
in this plot also the predictions for training and test set.

    ```{r, echo=TRUE}
    test_spl <- make_basis(test$speed)
    df_test <- as.data.frame(test_spl$X)
    test_pred <- predict(lm1, newdata = df_test)
    
    plot(train$speed, train$dist, col="blue", type="p",
         main="Train and Test Sets with Predictions",
         xlab="Distance",
         ylab="Speed",
         xlim=c(min(cars$speed), max(cars$speed))
         )
    points(test$speed, test$dist, col="green")
    lines(train$speed, train_pred, col="blue")
    lines(test$speed, test_pred, col="green")
    legend("topleft", legend=c("Train", "Test"), col=c("blue", "green"), pch=1)
    ```

    The predictions were about the same until past the 2nd knot 
    (`r round(spl$info$xquantiles[2], 3)`), where they then
    significantly diverge between the train and test sets.

4. Generate new speed data with \texttt{seq(0,30)}. Thus, we extend the range of 
the explanatory variable. Use the model derived above to obtain predictions, and 
present those in the plot from 3. (by first extending the x-range).

    ```{r, echo=TRUE}
    new_X <- seq(0,30)
    new_spl <- make_basis(new_X)
    df_new <- as.data.frame(new_spl$X)
    new_pred <- predict(lm1, newdata = df_new)
    
    plot(train$speed, train$dist, col="blue", type="p",
         main="Train and Test Sets with Predictions and Extended X",
         xlab="Distance",
         ylab="Speed",
         xlim=c(min(cars$speed), 30)
         )
    points(test$speed, test$dist, col="green")
    lines(train$speed, train_pred, col="blue")
    lines(test$speed, test_pred, col="green")
    lines(new_X, new_pred, col="orange")
    legend("topleft", legend=c("Train", "Test", "Extended"), 
           col=c("blue", "green", "orange"), 
           pch=1)
    ```
    
    The predictions on the extended dataset is in line with the others until
    the 1st knot (`r round(spl$info$xquantiles[1], 3)`), where it then gets
    driven down.
    
5. You might realize that the predictions from 4. are non-sense. The problem is 
that \texttt{lecturespl()} constructs the knots on the quantiles of the input 
variable. However, we should use the knots that have been constructed for the 
training data. Thus, modify the function accordingly, compute the predictions 
again, and visualize the new results.

    ```{r, echo=TRUE}
    lecturespl_mod1 <- function(x, nknots=2, M=4, splobj=NULL) {
      # nknots ... number of knots -> placed at regular quantiles
      # M ... M-1 is the degree of the polynomial
      n <- length(x)
      # X will not get an intercept column
      X <- matrix(NA,nrow=n,ncol=(M-1)+nknots)
      for (i in 1:(M-1)){ 
        X[,i] <- x^i 
      }
      
      # now the basis functions for the constraints:
      if (is.null(splobj)) {
        quant <- seq(0,1,1/(nknots+1))[c(2:(nknots+1))]
        qu <- quantile(x,quant)
      } else { # new addition
        quant <- splobj$quantiles
        qu <- splobj$xquantiles
      }

      for (i in M:(M+nknots-1)) {
        X[,i] <- ifelse(x-qu[i-M+1]<0,0,(x-qu[i-M+1])^(M-1))
      }
      list(X=X,quantiles=quant,xquantiles=qu)
    }
  
    make_basis_mod <- function(x, nknots=2, M=4, splobj=NULL) {
      out <- lecturespl_mod1(x, nknots, M, splobj) # adjusted arguments
      X <- out$X
      colnames(X) <- paste0("b", seq_len(ncol(X)))
      list(X = X, info = out)
      }
    
    # test
    test_spl <- make_basis_mod(x=test$speed, nknots=2, M=4, splobj=spl$info)
    df_test <- as.data.frame(test_spl$X)
    test_pred <- predict(lm1, newdata = df_test)
    
    # extended
    new_spl <- make_basis_mod(x=new_X, nknots=2, M=4, splobj=spl$info)
    df_new <- as.data.frame(new_spl$X)
    new_pred <- predict(lm1, newdata = df_new)
    
    # plotting
    plot(train$speed, train$dist, col="blue", type="p",
         main="Train and Test Sets with Predictions and Extended X",
         xlab="Distance",
         ylab="Speed",
         xlim=c(min(cars$speed), 30),
         )
    points(test$speed, test$dist, col="green")
    lines(train$speed, train_pred, col="blue")
    lines(test$speed, test_pred, col="green")
    lines(new_X, new_pred, col="orange")
    legend("topleft", legend=c("Train", "Test", "Extended"),
           col=c("blue", "green", "orange"),
           pch=1)
    
    # zoomed out
    plot(train$speed, train$dist, col="blue", type="p",
         main="Train and Test Sets with Predictions and Extended X",
         xlab="Distance",
         ylab="Speed",
         xlim=c(min(cars$speed), 30),
         ylim=c(-5, 550)
         )
    points(test$speed, test$dist, col="green")
    lines(train$speed, train_pred, col="blue")
    lines(test$speed, test_pred, col="green")
    lines(new_X, new_pred, col="orange")
    legend("topleft", legend=c("Train", "Test", "Extended"),
           col=c("blue", "green", "orange"),
           pch=1)
    ```

6. Another problem: Your predictions for low speed might look strange, as we 
expect for a speed of zero also a stipping distance of zero. Try to come up with 
a (simple) solution and present all results in one plot (training and test data, 
and the predictions for training, test and new x data).

    ```{r, echo=TRUE}
    # prevent negative response by using log transform
    lm1 <- lm(log(y) ~ ., data = df_train)
    
    # add column for intercept since missing in df_test
    X_test <- cbind("(Intercept)" = 1, df_new)
    
    # transpose multiplication of X_test with coef()
    log_pred <- as.numeric(as.matrix(X_test) %*% coef(lm1))
    new_pred <- exp(log_pred)
    
    # output new_pred
    new_pred # all values > 0
    
    # plotting
    plot(train$speed, train$dist, col="blue", type="p",
         main="Train and Test Sets with Predictions and Extended X",
         xlab="Distance",
         ylab="Speed",
         xlim=c(min(cars$speed), 30),
         )
    points(test$speed, test$dist, col="green")
    lines(train$speed, train_pred, col="blue")
    lines(test$speed, test_pred, col="green")
    lines(new_X, new_pred, col="orange")
    legend("topleft", legend=c("Train", "Test", "Extended"),
           col=c("blue", "green", "orange"),
           pch=1)
    
    # zoomed out
    plot(train$speed, train$dist, col="blue", type="p",
         main="Train and Test Sets with Predictions and Extended X",
         xlab="Distance",
         ylab="Speed",
         xlim=c(min(cars$speed), 30),
         ylim=c(-5, 550)
         )
    points(test$speed, test$dist, col="green")
    lines(train$speed, train_pred, col="blue")
    lines(test$speed, test_pred, col="green")
    lines(new_X, new_pred, col="orange")
    legend("topleft", legend=c("Train", "Test", "Extended"),
           col=c("blue", "green", "orange"),
           pch=1)
    ```

    The log transform fixes the issue of negative values for the response, though 
    now the model underpredicts when $x<25$ for the new x values compared to the
    train and test sets. When $x>25$, the log-transformed model predicts much 
    higher values compared to in task 5.

</div></pre>